Stack Safety as a Security Property

### 2020-07-03 main.tex
Fixes, nits, and variations

First of all, we are going to ignore the heap. We are mainly concerned
with stack interactions so that will be the main focus. \leo{We're
  going to come back to that at the end. Will we?} We are then going
to abstract away low-level details of memory layout and present a
machine that is word-addressed, with instructions starting from
location 0, a dedicated output at location 99, and lay the stack from
location 100 upwards. \leo{Explain more? How to connect to realistic
  one? Observations? DMA? Interrupt handlers?}
%
A sketch of the memory layout at that point can be found in the top of
Figure~\ref{fig:running-trace-b}: instructions for {\tt main} begin at
0, the dedicated output is empty, while the (uninitialized) stack
begins at 100. Directly below the layout, we summarize the parts of
the register file that are relevant to the example.
%
The rest of Figure~\ref{fig:running-trace-b} details the execution of the
program, showing the machine state at call and return points, as well
as describing the full effect of each instruction in detail.
\leo{Expand upon this?}

\bcp{ Comments on the example:
\begin{itemize}
\item We can simplify a bit by renaming $f$ to $\mathit{main}$ and
relocating its stack to location 1000.  And maybe move its instructions to
start at 0.  (We can comment that on a real RISC-V this is where interrupt
handlers go, but POPL readers will not care too much.)  We should streamline
the example as much as possible.
%
\item We should also emphasize, again, that there is no heap (if there were,
we might put it above the stack or between the instructions and the stack).
%
\item We will also need to be careful about fonts ($f$ vs {\texttt f}),
capitalization of register names (sp vs SP), etc. Maybe typeset instructions
in small caps?  Maybe typeset everything that is part of a concrete machine
in small caps??
%
\item The colors are helpful in making the big page of instruction traces easier
on the eye; maybe they would be even more helpful if
corresponding colors appeared elsewhere in the example (on the instructions
themselves, e.g.).  But we need to be careful to keep it readable -- maybe
they should be background colors, like you've done, rather than text colors.
\item The first column of numbers is redundant with the second column (PC).
Suggest deleting the first.
\item Asking readers to wade through all this is asking quite a bit!  We
need to make sure we are going to get sufficient value out of the example
later (I think we can, but we should really convince ourselves) to make it
worthwhile.
\item Use ``='' instead of ``:'' in the ``summaries'' below the pictures.
And put a bit more space between the components.
\item Create a new column that just literally pronounces each instruction in
English, so people don't have to read RISC-V opcodes.
\item As discussed, we should move this to before section 2, we should
extend it with a bit more at the end so we can talk about the lazy policy
with the same example, we should add
in some bad variants (doing it here replaces the ``Examples'' from the
Discussion section), and we should consider discussing the attacker model 
here (we need to discuss it someplace in any case!)
\end{itemize}
}

@@

\setcounter{pcctr}{27}
% H - Body
\vspace*{0.2em}
\[
\begin{array}{c|c|c|c|c}
  \row{\lw ~ r_4 ~ \rsp ~ 0}{r_4 \leftarrow [\rsp]}{r_4 \leftarrow 18}
      {Load return value to $r_4$}
  \row{\sw ~ \rsp ~ r_4 ~ \negate 3}{[\rsp - 3] \leftarrow r_4}{[105] \leftarrow 18}
      {Set {\tt w} to 18}
  \row{\addi ~ \rsp ~ \rsp ~ \negate 2}{\rsp \leftarrow \rsp - 2}{\rsp \leftarrow 106}
      {Deallocate {\tt h}'s call frame}
  \row{\lw ~ r_5 ~ \rsp ~ \negate 6}{r_5 \leftarrow [\rsp - 4]}{r_5 \leftarrow 0}
      {Load argument {\tt z} $r_5$}
  \row{\sw ~ \rout ~ r_5 ~ 0}{[\rout] \leftarrow r_5}{[99] \leftarrow 0}
      {Print {\tt z}}      
  \row{\add ~ r_4 ~ r_4 ~ r_5}{r_4 \leftarrow r_4 + r_5}{r_4 \leftarrow 18}
      {Add $r_4$ and $r_5$}
  \row{\sw ~ \rsp ~ r_4 ~ \negate 3}{[\rsp - 3] \leftarrow r_4}{[103] \leftarrow 18}
      {Store $r_4$ as ther return of {\tt g}}
  \row{\lw ~ \rra ~ \rsp ~ \negate 2}{\rra \leftarrow [\rsp - 2]}{\rra \leftarrow 8}
      {Load return address}
  \row{\addi ~ \rsp ~ \rsp ~ \negate 2}{\rsp \leftarrow \rsp - 2}{\rsp \leftarrow 104}
      {Deallocate {\tt g}'s local state}
  \row{\jalr ~ \rra ~ \rra ~ 0}{??}{PC \leftarrow 8}{Return to {\tt main}}
\end{array}
\]
% Return to main
% Memory Layout
\begin{center}
\MemoryLabel{4em}{2em}{0}
\MemoryLabel{21em}{2em}{99}
\MemoryLabel{43.5em}{2em}{\SP}
\memory{3}{gray}[{\makebox[0pt]{Instructions}}]%
\hspace*{3pt}
$\cdots$
\memory{1}{white}[{\makebox[0pt]{Output}}]%
\memory{2}{green}[{\makebox[0pt]{${\texttt main}_{priv}$}}]%
\memory{2}{green}[{\makebox[0pt]{{\tt g}'s call frame }}]%
\memory{6}{gray}%[{\makebox[0pt]{Uninitialized}}]
~$\cdots$
\MemoryLabel{-51em}{0.75em}{0}
\MemoryLabel{-47em}{0.75em}{42}
\MemoryLabel{-43em}{0.75em}{0}
\MemoryLabel{-39em}{0.75em}{0}
\MemoryLabel{-35em}{0.75em}{18}
\\
\end{center}
% Register File
\summary{8}{0}{\thepcctr}{104}{18}{0}
%\vspace*{0.2em}
\setcounter{pcctr}{8}
\[
\begin{array}{c|c|c|c|c}
  \row{\lw ~ \rsp ~ r_4 ~ \negate 1}{r_4 \leftarrow [\rsp - 1]}{r_4 \leftarrow 18}
      {Load return value to $r_4$}
  \row{\sw ~ \rsp ~ r_4 ~ \negate 3}{[\rsp - 3] \leftarrow r_4}{[101] \leftarrow 18}
      {Set {\tt y} to the result}
  \row{\addi ~ \rsp ~ \rsp ~ \negate 4}{\rsp \leftarrow \rsp - 2}{\rsp \leftarrow 102}
      {Deallocate {\tt g}'s call frame}
  \row{\lw ~ \rsp ~ r_5 ~ \negate 2}{r_5 \leftarrow [\rsp - 2]}{r_5 \leftarrow 42}
      {Load {\tt x} to $r_5$}
  \row{\add ~ r_4 ~ r_4 ~ r_5}{r_4 \leftarrow r_4 + r_5}{r_4 \leftarrow 60}
      {Add {\tt x} and {\tt y}}
  \row{\sw ~ \rout ~ r_4 ~ 0}{[\rout] \leftarrow r_4}{[99] \leftarrow 60}
      {Print {\tt x + y}}
\end{array}
\]
\vspace*{0.2em}
% Return to main
% Memory Layout
\begin{center}
\MemoryLabel{4em}{2em}{0}
\MemoryLabel{21em}{2em}{99}
\MemoryLabel{31.5em}{2em}{\SP}
\memory{3}{gray}[{\makebox[0pt]{Instructions}}]%
\hspace*{3pt}
$\cdots$
\memory{1}{white}[{\makebox[0pt]{Output}}]%
\memory{2}{green}[{\makebox[0pt]{${\texttt main}_{priv}$}}]%
\memory{8}{gray}%[{\makebox[0pt]{Uninitialized}}]
~$\cdots$
\MemoryLabel{-51em}{0.75em}{60}
\MemoryLabel{-47em}{0.75em}{60}
\MemoryLabel{-43em}{0.75em}{42}
\\
\end{center}
% Starting Register File
\summary{\thepcctr}{0}{37}{102}{60}{42}
\caption{Detailed Execution Trace}
\label{fig:running-trace-b}
\end{figure}

%\end{minipage}
%\begin{minipage}{0.7\textwidth}
%\newcommand*{\instr}[3]{ & #1: & #2 & ; \text{\em #3}\\}
%\newcommand*{\clabel}[1]{#1: & & &\\}
%\[
%\begin{array}{llll}
%  \clabel{TOP}
%  \instr{100}{\addi ~ \rsp ~ \rsp ~ 8}{allocate 2 words in stack}
%  \instr{104}{\sw ~ \rsp ~ r_0 ~ -4}{store 0 as an argument}
%  \instr{108}{\sw ~ \rsp ~ r_0 ~ 0}{zero out the return location}
%  \instr{112}{\jal ~ \rra ~ F}{call F and store the return address}
%  & & & \\
%  & & & \\
%  \clabel{F}
%  \instr{140}{\addi ~ \rsp ~ \rsp ~ 4}{allocate for the return address}
%  \instr{144}{\sw ~ \rsp ~ \rra ~ 0}{store the return address to sp}
%  \instr{148}{\addi ~ \rsp \rsp ~ 8}{allocate for nested arg + ret}
%  \instr{152}{\addi ~ r_4 ~ r_0 ~ 17}{put 17 in register $r_4$}
%  \instr{156}{\sw ~ \rsp ~ r_4 ~ -4}{store 17 as the argument to G}
%  \instr{160}{\sw ~ \rsp ~ r_0 ~ 0}{zero out the return location}
%  \instr{164}{\jal ~ \rra ~ G}{call G and store the return address}
%  \instr{168}{\lw ~ r_4 ~ \rsp}{load the return value of G to $r_4$}
%  \instr{172}{\lw ~ r_5 ~ \rsp -12}{load the argument of F to $r_5$}
%  \instr{176}{\add ~ r_4 ~ r_4 ~ r_5}{add $r_4$ and $r_5$}
%  \instr{172}{\sw ~ \rsp ~ r_4 ~ -8}{store $r_4$ as ther return of F}
%  \instr{176}{\addi ~ \rsp ~ \rsp ~ -8}{deallocate local state}
%  \instr{180}{\lw ~ \rra ~ \rsp ~ -4}{load the return address from the stack pointer}
%  \instr{184}{\addi ~ \rsp ~ \rsp ~ -4}{restore the stack pointer of F}
%  \instr{188}{\jalr ~ \rra ~ \rra ~ 0}{return to TOP}
%  & & & \\
%  & & & \\
%  \clabel{G}
%  \instr{200}{\addi ~ \rsp ~ \rsp ~ 4}{allocate for the return address}
%  \instr{204}{\sw ~ \rsp ~ \rra ~ 0}{store the return address to sp}
%  \instr{208}{\lw ~ r_4 ~ \rsp ~ -12}{load the argument of G to $r_4$}
%  \instr{212}{\addi ~ r_4 ~ r_4 ~ 1}{increment $r_4$ by 1}
%  \instr{216}{\sw ~ \rsp ~ r_4 ~ -8}{store $r_4$ as the return value}
%  \instr{220}{\lw ~ \rra ~ \rsp ~ -4}{load the return address from the stack pointer}
%  \instr{224}{\addi ~ \rsp ~ \rsp ~ -4}{restore the stack pointer of F}
%  \instr{228}{\jalr ~ \rra ~ \rra ~ 0}{return to F}
%\end{array}
%\]
%\end{minipage}
%\label{fig:running-program}
%\end{figure}

%  sw r2 42 -8; store 42 in a private location
%  sw r2 0 -4; store 0 as an argument
%  sw r2 0 0; zero out the return location
%  jal r1 F ; call F and store return address

%  0: & \addi ~ r_4 ~ r_0 ~ 40 &
%     ; ~ \text{\em put $40$ in register $r_4$}\\
%  4: & \sw ~ r_4 ~ r_4 ~ 1000 &
%     ; ~ \text{\em store $40$ in location $1040$}\\
%  8: & \jal ~ r_1 ~ 20        &
%     ; ~ \text{\em jump to $20$, store return addres in $r_1$}\\
%  12: & & \\
%  16: & & \\
%  20: & \sw ~ r_2 ~ r_1 ~ 4 &
%     ; ~ \text{\em store the return address to the $\SP + 4$}\\
%  24: & \addi ~ r_2 ~ r_2 ~ 8 &
%     ; ~ \text{\em increase the $\SP$ by 8}\\
%\end{array}
%\]
%
%  , (SW sp ra 4  , tagH1)
%  , (ADDI sp sp 8, tagH2)
%  , (LW 3 3 1000, instrTag)
%  ]

>>>>

@@

\setcounter{pcctr}{27}
% H - Body
\vspace*{0.2em}
\[
\begin{array}{c|c|c|c|c}
  \row{\lw ~ r_4 ~ \rsp ~ 0}{r_4 \leftarrow [\rsp]}{r_4 \leftarrow 18}
      {Load return value to $r_4$}
  \row{\sw ~ \rsp ~ r_4 ~ \negate 3}{[\rsp - 3] \leftarrow r_4}{[105] \leftarrow 18}
      {Set {\tt w} to 18}
  \row{\addi ~ \rsp ~ \rsp ~ \negate 2}{\rsp \leftarrow \rsp - 2}{\rsp \leftarrow 106}
      {Deallocate {\tt h}'s call frame}
  \row{\lw ~ r_5 ~ \rsp ~ \negate 6}{r_5 \leftarrow [\rsp - 6]}{r_5 \leftarrow 0}
      {Load argument {\tt z} $r_5$}
  \row{\sw ~ \rout ~ r_5 ~ 0}{[\rout] \leftarrow r_5}{[99] \leftarrow 0}
      {Print {\tt z}}      
  \row{\add ~ r_4 ~ r_4 ~ r_5}{r_4 \leftarrow r_4 + r_5}{r_4 \leftarrow 18}
      {Add $r_4$ and $r_5$}
  \row{\sw ~ \rsp ~ r_4 ~ \negate 3}{[\rsp - 3] \leftarrow r_4}{[103] \leftarrow 18}
      {Store $r_4$ as ther return of {\tt g}}
  \row{\lw ~ \rra ~ \rsp ~ \negate 2}{\rra \leftarrow [\rsp - 2]}{\rra \leftarrow 8}
      {Load return address}
  \row{\addi ~ \rsp ~ \rsp ~ \negate 2}{\rsp \leftarrow \rsp - 2}{\rsp \leftarrow 104}
      {Deallocate {\tt g}'s local state}
  \row{\jalr ~ \rra ~ \rra ~ 0}{??}{PC \leftarrow 8}{Return to {\tt main}}
\end{array}
\]
% Return to main
% Memory Layout
\begin{center}
\MemoryLabel{4em}{2em}{0}
\MemoryLabel{21em}{2em}{99}
\MemoryLabel{43.5em}{2em}{\SP}
\memory{3}{gray}[{\makebox[0pt]{Instructions}}]%
\hspace*{3pt}
$\cdots$
\memory{1}{white}[{\makebox[0pt]{Output}}]%
\memory{2}{green}[{\makebox[0pt]{${\texttt main}_{priv}$}}]%
\memory{2}{green}[{\makebox[0pt]{{\tt g}'s call frame }}]%
\memory{6}{gray}%[{\makebox[0pt]{Uninitialized}}]
~$\cdots$
\MemoryLabel{-51em}{0.75em}{0}
\MemoryLabel{-47em}{0.75em}{42}
\MemoryLabel{-43em}{0.75em}{0}
\MemoryLabel{-39em}{0.75em}{0}
\MemoryLabel{-35em}{0.75em}{18}
\\
\end{center}
% Register File
\summary{8}{0}{\thepcctr}{104}{18}{0}
%\vspace*{0.2em}
\setcounter{pcctr}{8}
\[
\begin{array}{c|c|c|c|c}
  \row{\lw ~ \rsp ~ r_4 ~ \negate 1}{r_4 \leftarrow [\rsp - 1]}{r_4 \leftarrow 18}
      {Load return value to $r_4$}
  \row{\sw ~ \rsp ~ r_4 ~ \negate 3}{[\rsp - 3] \leftarrow r_4}{[101] \leftarrow 18}
      {Set {\tt y} to the result}
  \row{\addi ~ \rsp ~ \rsp ~ \negate 4}{\rsp \leftarrow \rsp - 2}{\rsp \leftarrow 102}
      {Deallocate {\tt g}'s call frame}
  \row{\lw ~ \rsp ~ r_5 ~ \negate 2}{r_5 \leftarrow [\rsp - 2]}{r_5 \leftarrow 42}
      {Load {\tt x} to $r_5$}
  \row{\add ~ r_4 ~ r_4 ~ r_5}{r_4 \leftarrow r_4 + r_5}{r_4 \leftarrow 60}
      {Add {\tt x} and {\tt y}}
  \row{\sw ~ \rout ~ r_4 ~ 0}{[\rout] \leftarrow r_4}{[99] \leftarrow 60}
      {Print {\tt x + y}}
\end{array}
\]
\vspace*{0.2em}
% Return to main
% Memory Layout
\begin{center}
\MemoryLabel{4em}{2em}{0}
\MemoryLabel{21em}{2em}{99}
\MemoryLabel{31.5em}{2em}{\SP}
\memory{3}{gray}[{\makebox[0pt]{Instructions}}]%
\hspace*{3pt}
$\cdots$
\memory{1}{white}[{\makebox[0pt]{Output}}]%
\memory{2}{green}[{\makebox[0pt]{${\texttt main}_{priv}$}}]%
\memory{8}{gray}%[{\makebox[0pt]{Uninitialized}}]
~$\cdots$
\MemoryLabel{-51em}{0.75em}{60}
\MemoryLabel{-47em}{0.75em}{60}
\MemoryLabel{-43em}{0.75em}{42}
\\
\end{center}
% Starting Register File
\summary{\thepcctr}{0}{37}{102}{60}{42}
\caption{Detailed Execution Trace}
\label{fig:running-trace-b}
\end{figure}

\bcp{We should also emphasize, again, that there is no heap (if there were,
we might put it above the stack or between the instructions and the stack)}
First of all, we are going to ignore the heap. We are mainly concerned
with stack interactions so that will be the main focus. \leo{We're
  going to come back to that at the end. Will we?} We are then going
to abstract away low-level details of memory layout and present a
machine that is word-addressed, with instructions starting from
location 0, a dedicated output at location 99, and lay the stack from
location 100 upwards. \leo{Explain more? How to connect to realistic
  one? Observations? DMA? Interrupt handlers?}
%
A sketch of the memory layout at that point can be found in the top of
Figure~\ref{fig:running-trace-b}: instructions for {\tt main} begin at
0, the dedicated output is empty, while the (uninitialized) stack
begins at 100. Directly below the layout, we summarize the parts of
the register file that are relevant to the example.
%
The rest of Figure~\ref{fig:running-trace-b} details the execution of the
program, showing the machine state at call and return points, as well
as describing the full effect of each instruction in detail.
\leo{Expand upon this?}

\bcp{ Comments on the example:
\begin{itemize}
\item We will also need to be careful about fonts ($f$ vs {\texttt f}),
capitalization of register names (sp vs SP), etc. Maybe typeset instructions
in small caps?  Maybe typeset everything that is part of a concrete machine
in small caps??
%
\item The colors are helpful in making the big page of instruction traces easier
on the eye; maybe they would be even more helpful if
corresponding colors appeared elsewhere in the example (on the instructions
themselves, e.g.).  But we need to be careful to keep it readable -- maybe
they should be background colors, like you've done, rather than text colors.
\item Asking readers to wade through all this is asking quite a bit!  We
need to make sure we are going to get sufficient value out of the example
later (I think we can, but we should really convince ourselves) to make it
worthwhile.
\item Create a new column that just literally pronounces each instruction in
English, so people don't have to read RISC-V opcodes.
\item As discussed, we should move this to before section 2, we should
extend it with a bit more at the end so we can talk about the lazy policy
with the same example, we should add
in some bad variants (doing it here replaces the ``Examples'' from the
Discussion section), and we should consider discussing the attacker model 
here (we need to discuss it someplace in any case!)
\end{itemize}
}

\subsection{Attacker Model + Unsafe Variations}

\leo{TODO: Make attacker model more explicit} The execution trace
shown in Figure~\ref{fig:running-trace-b} is an example of a
``stack-safe'' execution: each process interacts only with the
registers, its local stack space and its call frame. But what could go
wrong if an attacker could hijack an arbitrary call and execute
arbitrary code? In the rest of this section we'll go over examples of
unsafe behaviors, before formally capturing what stack-safe means
formally and discussing enforcement techniques from the literature
in later sections.

\paragraph*{Bad write}

\[
44 : ~ \sw ~ \rsp ~ r_4 ~ \negate 5
\]

Instruction 44 currently is $\sw ~ \rsp ~ r_4 ~ \negate 2$, with the
effect of storing the contents of $r_4$ as the result of {\tt h}. If
instead it wrote to location $[\rsp - 5]$ then it would be overriding
private data (the return address) belonging to {\tt g}.

\paragraph*{Bad read}

\[
30 : ~ \lw ~ r_5 ~ \rsp ~ \negate 8
\]

Instruction 30 currently is $\lw ~ r_5 ~ \rsp ~ \negate 6$, with the
effect of loading the argument of {\tt g} to $r_4$. If instead it
loaded the contents of location $[\rsp - 8]$, it would obtain access
to {\tt main}'s private information. Worse, since the next instruction
outputs the contents of $r_5$ that would immediately be made public.

\paragraph*{Control flow}

\leo{What should we do here? Overwrite our own ra and return?}
